Historian: accurate reconstruction of ancestral sequences and evolutionary rates

### 2016-10-30 doc/main.tex
Updated

\abstract{
{\bf Motivation.}
Reconstruction of ancestral sequence histories,
and estimation of parameters like indel rates,
are improved by using explicit evolutionary models
and summing over uncertain alignments.
%The best alignment tools for ancestral reconstruction
%are not necessarily those that agree most closely with benchmark datasets based on 3D structure alignments.
The previous best tool for this purpose was ProtPal,
but this tool was too slow for practical use.
{\bf Results.}
Historian combines an efficient reimplementation of the ProtPal algorithm with performance-improving heuristics from other alignment tools.
Simulation results on fidelity of ancestral sequence reconstruction,
along with evaluations on the structurally-informed dataset BAliBase 3.0,
recommend Historian for evolutionary applications.
{\bf Availability and Implementation.}
Historian is available at \url{https://github.com/ihh/indelhistorian} under the Creative Commons Attribution 3.0 US license.
{\bf Contact.}
Ian Holmes {\tt ihholmes+indelhistorian@gmail.com}.
{\bf Supplementary Information.}
None.
}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{sim/results.pdf}
  \caption{
    Results of the simulation benchmark.
    For each method evaluated, the distribution of the ratio of inferred to true
    indel rate is shown, along with the mean and root-mean-squared error for this
    ratio (which, for a perfect rate inference, would be equal to 1).
  }
\end{figure}

@@

As noted in Table~2, some of the results were taken from the Muscle website, {\tt drive5.com/bench/} \citep{Edgar2010}.
To make a direct comparison of runtimes, we re-ran the benchmarks for Prank and Muscle on the same machine as the Historian benchmark.
These runtimes are summarized in Table~3: Historian is an order of magnitude slower than Muscle, but an order of magnitude faster than Prank.
Re-running Prank and Muscle resulted in a slight improvement for Prank
and a slight deterioration for Muscle compared to the {\tt drive5.com} data,
presumably due to versioning issues
(the version of Muscle available for download 3.8.31, whereas the data
reported are for version 4.0; conversely, a more recent version of Prank is available
than the one benchmarked on the Muscle website).
Table~2 reflects our results, where available, and the {\tt drive5.com} results in other cases.

@@

Historian is available at \url{github.com/ihh/indelhistorian} under the CC BY 3.0 US license. It is written in C++11, compiles on a POSIX system with Clang (v.6.1.0) and requires several free libaries (libgsl, libz and Boost).
The simulation data reported in this paper are available at \url{github.com/ihh/gp120sim}.

>>>>

\abstract{
{\bf Motivation.}
Reconstruction of ancestral sequence histories,
and estimation of parameters like indel rates,
are improved by using explicit evolutionary models
and summing over uncertain alignments.
%The best alignment tools for ancestral reconstruction
%are not necessarily those that agree most closely with benchmark datasets based on 3D structure alignments.
The previous best tool for this purpose was ProtPal,
but this tool was too slow for practical use.
{\bf Results.}
Historian combines an efficient reimplementation of the ProtPal algorithm with performance-improving heuristics from other alignment tools.
Simulation results on fidelity of rate estimation via ancestral reconstruction,
along with evaluations on the structurally-informed alignment dataset BAliBase 3.0,
recommend Historian over other alignment tools for evolutionary applications.
{\bf Availability and Implementation.}
Historian is available at \url{https://github.com/ihh/indelhistorian} under the Creative Commons Attribution 3.0 US license.
{\bf Contact.}
Ian Holmes {\tt ihholmes+historian@gmail.com}.
{\bf Supplementary Information.}
None.
}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{sim/results.pdf}
  \caption{
    Results of the simulation benchmark.
    For each method evaluated, the distribution of the ratio of inferred to true
    indel rate is shown, along with the mean and root-mean-squared error for this
    ratio (which, for a perfect rate inference, would be equal to 1).
    Tool versions are as shown in Table~1.
  }
\end{figure}

@@

As noted in Table~2, some of the results were taken from the Muscle website, {\tt drive5.com/bench/} \citep{Edgar2010}.
To make a direct comparison of runtimes, we re-ran the benchmarks for Prank and Muscle on the same CPU as the Historian benchmark (Intel Xeon 3.20GHz).
The runtimes are summarized in Table~3: Historian is an order of magnitude slower than Muscle, but an order of magnitude faster than Prank.
Re-running Prank and Muscle resulted in a slight improvement for Prank
and a slight deterioration for Muscle compared to the {\tt drive5.com} data,
presumably due to versioning issues
(the version of Muscle available for download 3.8.31, whereas the data
reported are for version 4.0; conversely, a more recent version of Prank is available
than the one benchmarked on the Muscle website).
Table~2 reflects our results, where available, and the {\tt drive5.com} results in other cases.

@@

Historian is available at \url{github.com/ihh/indelhistorian} under the CC BY 3.0 US license.
Precompiled binaries for Linux or OSX are available.
The program may also be compiled from the C++11 source
on a POSIX system with Clang (v.6.1.0) and several free libaries (libgsl, libz and Boost).
Makefiles to reproduce the analyses reported in this paper
are included in the repository, and the simulation data
are available at \url{github.com/ihh/gp120sim}.
